Correlation tests

### 2025-04-16 analysis/manuscript.tex
Some minor language fixes

\section*{Abstract}
{}
{\textbf{Keywords}: hybrid open access, transformative agreements, scholarly publishing, big deals, bibliometrics}

@@

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, allowing authors from participating
institutions to publish open access in these journals, while providing
reading access to the entire portfolio. Library consortia and funders
have begun coordinating how these agreements are negotiated, managed and
evaluated. At the same time, efforts are underway to improve
transparency through open data on contracts and payments. However, this
information is still only partially available, which limits large-scale
studies of the impact of transformative agreements on the transition of
hybrid journals to full open access. By comparing open metadata from
hoaddata, an openly available dataset on hybrid open access based on
metadata from Crossref, OpenAlex and the cOAlition S Journal Checker
Tool, with the established bibliometric databases Scopus and Web of
Science, this study aims to provide a robust estimate of the impact of
transformative agreements on hybrid open access. This comparative
approach informs bibliometric research and practice by highlighting the
strengths and weaknesses of each data source individually and
collectively.

Since their initial proposal by the Max Planck Digital Library (Schimmer
et al., 2015), transformative agreements have led to a rapid growth in
the number of negotiated agreements and open access articles (Jahn,
2025; McCabe \& Mueller-Langer, 2024; Rothfritz et al., 2024). Before,
only a small proportion of articles in hybrid journals were openly
available (Björk, 2012; Laakso \& Björk, 2016; Piwowar et al., 2018),
with publications from only a few European countries with dedicated
funding policies and the high-energy physics SCOAP\^{}3 consortium
standing out (Huang et al., 2020; Kohls \& Mele, 2018; Pinfield, 2015;
Robinson-Garcia et al., 2020). Related coordination activities, in
particular through the OA2020 and ESAC (Efficiency and Standards for
Article Charges) initiatives, have also led to improved workflows
between library consortia and publishers (Campbell et al., 2022;
Geschuhn \& Stone, 2017). However, critics have raised concerns about
these agreements, particularly regarding market concentration (Butler et
al., 2023; Shu \& Larivière, 2023), the failure of most hybrid journals
to convert to full open access (Kiley, 2024; Momeni et al., 2021),
reduced competition and incentives (McCabe \& Mueller-Langer, 2024;
Schmal, 2024), and continuation of publication fees that widen gaps
between well-resourced and under-resourced institutions (Babini et al.,
2022; Ross-Hellauer et al., 2022)

While consortia evaluations confirm the growth of open access through
transformative agreements and acknowledge increasing coordination, they
reach mixed policy conclusions. For instance, the British Jisc
evaluation (Brayman et al., 2024) found that the consortium's
transformative agreements had a significant impact on national open
access growth but a limited effect on the global open access transition,
suggesting a need to reevaluate their strategy. Similar policy
reconsiderations emerged in Norway and Sweden (Holden et al., 2023;
Widding, 2024). Meanwhile, the funder initiative cOAlition S recommended
ending financial support for these agreements in 2024, but still
considers hybrid open access to be compliant with funders' open access
policies. Conversely, the German DEAL consortium extended agreements
with major publishers until 2028, while also outside of Europe more and
more transformative agreements are being put in place (for instance in
Latin America, as described in Muñoz-Vélez et al. (2024)). Responding to
criticism, the Max Planck Digital Library recently outlined strategies
to evolve the original model, focusing on achieving full open access
while controlling costs and addressing identified weaknesses (Dér, 2025)

Despite controversies, transformative agreements can benefit the
landscape of open research information as library consortia negotiate
for comprehensive open metadata, helping avoid data analytics becoming a
commodity controlled by publishers (Aspesi \& Brand, 2020; McCabe \&
Mueller-Langer, 2024). For instance, ESAC recommends using Crossref
(Hendricks et al., 2020) to share metadata supporting workflows
including open access licenses and funding information (Geschuhn \&
Stone, 2017). Simultaneously, the bibliometric community pushes for more
open metadata through this publisher-driven DOI registration platform
and continuously monitors progress (N. J. van Eck \& Waltman, 2022).

Many open scholarly data services that build on Crossref benefit from
these developments. A prominent example is OpenAlex, which has gained
attention for its inclusivity, openness and comprehensive coverage of
publication metadata (Priem et al., 2022). In a recent evaluation,
Céspedes et al. (2025) reviewed the current evidence base on the
strengths and weaknesses of OpenAlex. They highlight both its increasing
use in bibliometric research and its adoption by research organisations.
Comparative studies have generally found OpenAlex to be a reliable
alternative to commercial databases (Alperin et al., 2024; Culbert et
al., 2024), although data quality issues occurred for key metadata
fields needed for open access analysis, including evidence of open
access status itself (Jahn et al., 2023; Simard et al., 2024), as well
as authorships (Culbert et al., 2024), affiliations (Zhang et al.,
2024), author roles (Jahn, 2025) and document types used to distinguish
between different article types (Haupka et al., 2024). However, these
studies highlight the rapid improvements and the new possibilities of
combining open metadata and the curation efforts of OpenAlex, in
conjunction with the growing awareness of the importance of open
metadata, as demonstrated by the Barcelona Declaration on Open Research
Information.

As a consequence, there is growing interest in using open bibliometric
databases such as Crossref and OpenAlex to assess the impact of
transformative agreements. Similar to established proprietary
bibliometric databases such as Scopus and Web of Science, these data
sources lack specific data points, particularly about open access
payments, which are typically not shared (Haustein et al., 2024).
However, bibliometric data can be enriched with open metadata about the
journals and institutions involved in transformative agreements (Jahn,
2025). Jonge et al. (2025) demonstrated this potential by combining data
from the cOAlition S Journal Checker tool and OpenAlex, achieving 89\%
accuracy when validated against Dutch research information. While open
data approaches show promise, understanding how these results compare
when the same methodology is applied to proprietary databases such as
Web of Science and Scopus remains unexplored.

This study compares open and proprietary data sources for evaluating
transformative agreements. It adapts the approach from Jahn (2025),
which combines data from the cOAlition S Journal Checker tool, Crossref
and OpenAlex, and applies the same methods to Web of Science and Scopus
data. This comparative approach allows the evaluation of different
bibliometric data sources for monitoring transformative agreements,
while assessing their respective strengths and limitations.
Specifically, this study compares the indexing of more than 13,000
hybrid journals across these sources, and examines whether consistent
results can be obtained in terms of open access adoption between 2019
and 2023, analysed by publisher and country. By comparing results across
these data sources, this large-scale study provides a more robust
understanding of transformative agreements' impact on the transition to
full open access. It also reveals the comparative strengths and
weaknesses of different bibliometric databases in measuring scholarly
publishing's shift toward openness

@@

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex.\footnote{\url{https://subugoe.github.io/scholcomm_analytics/data.html}}
The workflow has run regularly to fetch updates from these data sources
since 2022. The package version used in this study is 0.3, containing
data from the Crossref 2024-08 dump provided to Crossref Metadata Plus
subscribers and the OpenAlex 2024-08-29 monthly dump. It covers
agreements collected between July 2021 to July 2024 from the JCT. This
version including the computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is a selective, base
research-focused database (Stahlschmidt \& Stephen, 2022; Visser et al.,
2021). The collections considered in this study were the Science
Citation Index Expanded (SCIE), the Social Sciences Citation Index
(SSCI) and the Arts \& Humanities Citation Index (AHCI).

These collections provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall (Piwowar et al.,
2018), the same provider as OpenAlex. However, Web of Science lacks
information about journals and articles under transformative agreements.

For programmatic access to article-level data, this study used the
database of the Kompetenznetzwerk Bibliometrie (KB) in Germany. The KB
processes raw XML data provided by Clarivate Analytics, which is
ingested into an in-house PostgreSQL database under a uniform schema. To
support reproducibility, KB maintains annual snapshots of the database.
Accordingly, this study used the annual snapshot from April 2024
(wos\_b\_202404), which is considered to cover almost the entire
previous publication year (Schmidt et al., 2024).

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to Web of Science, Scopus is selective with regard to the
journals it indexes. However, its journal coverage is much broader than
that of the Web of Science collections considered in this study, as it
also indexes a wider range of applied research journals (Singh et al.,
2021; Stahlschmidt \& Stephen, 2022; Visser et al., 2021). With detailed
metadata about article types, open access status information derived
from Unpaywall, author roles, and disambiguated affiliations, Scopus
also contains important data to assess open access uptake, although no
direct information regarding transformative agreements was available at
the time of the study.

This study used the Scopus annual snapshot of April 2024 as provided by
the KB (scp\_b\_202404). The same KB curation effort as for the Web of
Science was applied to the Scopus (Schmidt et al., 2024).

@@

\subsubsection{Determining hybrid journal publication
volume}\label{determining-hybrid-journal-publication-volume}

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
The JCT journal data were enriched with ISSN variants linked to an
ISSN-L. To identify hybrid journals, a comprehensive exclusion of fully
open access journals was performed using multiple journal lists
including the Directory of Open Access Journals (DOAJ). The resulting
hybrid journal data were made available via hoaddata and used to
determine the publication volume for each database independently.

hoaddata relies on Crossref for obtaining journal publication volume and
open access status through Creative Commons (CC) licence information
relative to the published version (``version of record''). The article
metadata included DOIs, publication dates, open access information as
well as author roles and affiliations. Publication years were determined
using the earliest known date of publication in a journal. In hoaddata,
this corresponded to Crossref's issued date. For Web of Science and
Scopus, the earliest publication date was used where available, with
Scopus dates specifically determined by the KB through version tracking
of the raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews (Borrego et al., 2021). Because of limited information on these
document types in open scholarly data (Haupka et al., 2024), hoaddata
used an extended version of Unpaywall's paratext recognition approach to
exclude non-scholarly content (Piwowar et al., 2018). To exclude
conference supplements, which are also often not covered by
transformative agreements, only articles published in regular issues,
indicated by numerical pagination, were considered. For Web of Science
and Scopus, their established, mainly accurate document type
classifications (Donner, 2017; Maisano et al., 2025) were used to
identify original research articles and reviews, referred to as original
articles throughout this study.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a CC license on publisher platforms. While
hoaddata obtained this information from Crossref license metadata, Web
of Science and Scopus relied on Unpaywall as evidence source. Unpaywall
also uses Crossref license metadata, but supplements them by parsing
publisher websites directly, addressing cases where publishers do not
provide machine-readable CC license information (Piwowar et al., 2018).
This additional parsing remains necessary despite transformative
agreement workflows recommend to deposit CC licenses during DOI
registration (Geschuhn \& Stone, 2017). Both Web of Science\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}
and Scopus\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}}
defined hybrid open access consistently as content available under CC
licenses on publisher platforms, distinguishing it from bronze open
access that lack such explicit license information, or use
publisher-specific licenses (Piwowar et al., 2018).

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To improve data
retrieval, JCT institution data was enriched with ROR-IDs from
associated institutions, such as university hospitals or institutes of
large research organisations such as the Max Planck Society, according
to OpenAlex' institution entity. To handle different address variants,
database-specific affiliation identifiers were used: ROR-IDs from
OpenAlex for hoaddata, affiliation enhanced names for Web of Science,
and Scopus Affiliation Identifier. Additionally, ISO country codes were
retrieved for each author's address to compile country-level statistics.

Because neither Web of Science nor Scopus supported ROR-ID at the time
of the data retrieval, the institution identifier used by the JCT, a
two-step matching process was implemented to harmonise affiliation data.
First, 2,782,540 articles from 6,457 institutions with ROR-IDs in the
JCT data since 2017 (according to hoaddata) were processed to map first
authors' ROR-IDs to corresponding proprietary affiliation identifier in
Web of Science and Scopus using DOI matching. Then, an algorithm
selected the most frequent ROR ID and proprietary identifier pairs to
handle multiple affiliations and organisational hierarchy differences.

This process linked 6,375 ROR-IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR-IDs to 2,422 enhanced affiliation strings in Web of Science.
Quality evaluation through random sampling of 50 pairs revealed an error
rate of 22\% for Web of Science (11 mismatches) and 6\% for Scopus (3
mismatches). Upon inspection, these mismatches primarily occurred with
less-represented institutions having only a few publications, introduced
through multiple affiliations of single authors. The difference between
databases suggests that Scopus' affiliation control aligns more closely
with ROR than that of Web of Science.

@@

Based on these compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain ROR-IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions according to
the JCT. The matching also took into account the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (Jonge et al., 2025), applied to publications
funded by the Dutch Research Council (NWO) and validated against
internal invoicing data, confirmed that such matching can accurately
identify most articles under transformative agreements.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023. It shows that the
majority of hybrid journals published at least one original research
article or review, collectively marked as ``original article'' in the
following, during the five-year period. These journals provided the
basis for the subsequent calculation of article-level indicators.

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, because Unpaywall only collects information on open access
status for articles with a DOI. A closer look at original articles with
affiliation data, and in line with related research (Zhang et al.,
2024), reveals a lack of affiliation data in the case of OpenAlex, the
affiliation data source used by hoaddata, compared to Web of Science and
Scopus. In particular, only about two-thirds of the articles examined
provided corresponding author affiliations. For first authors, the
proportion was 89\%. At the time of writing, OpenAlex disclosed limited
coverage of corresponding authorship data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

@@

Throughout this largely automated data collection and analysis process,
Tidyverse tools (Wickham et al., 2019) for the R programming language (R
Core Team, 2024) were used. Rank correlations were calculated using the
Hmisc package (Harrell Jr, 2003). The source code analysis, including
all queries used to obtain the data, is available on GitHub (add link).
An interactive supplement exploring correlations between the data
sources examined by country and publisher is available on HuggingFace
Spaces: https://huggingface.co/spaces/najkoja/hoa\_replication.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals and articles are visualised as an UpSet graph (Krassowski,
2020; Lex et al., 2014). The analysis included hybrid journals that
published at least one open access article between 2019 and 2023, based
on open access status information from each database. Only original
articles were considered in the analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were only found in Scopus. Of these, 354 were also available in
Web of Science. Upon inspection, this group of hybrid journals
exclusively covered in the proprietary data sources mainly represented
hybrid journals for which no open access evidence could be retrieved
from Crossref, the open access evidence source for hoaddata.

>>>>

\section*{Abstract}
{This study compares open metadata from hoaddata, an openly available dataset based on Crossref, OpenAlex and the cOAlition S Journal Checker Tool, with proprietary bibliometric databases Scopus and Web of Science to estimate the impact of transformative agreements on hybrid open access publishing. Analysing over 13,000 hybrid journals between 2019-2023, the research found substantial growth in open access due to these agreements, although most articles remain paywalled. The results were consistent across all three data sources, showing strong correlations in country-level metrics despite differences in journal coverage and metadata availability. By 2023, transformative agreements enabled the majority of hybrid open access articles, with particularly high adoption in European countries. The analysis revealed strong alignment between first and corresponding authorship when measuring agreement uptake by publisher and country. This comparative approach supports the use of open metadata for large-scale hybrid open access studies, while using multiple data sources together provides a more robust understanding of hybrid open access adoption than any single database can offer, overcoming individual limitations in coverage and metadata quality.}

{\textbf{Keywords}: bibliometric data sources, open metadata, hybrid open access, transformative agreements, curative bibliometrics}

@@

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, allowing authors from participating
institutions to publish open access in these journals while providing
reading access to the entire portfolio. Library consortia and funders
have begun to coordinate how these agreements are negotiated, managed
and evaluated. At the same time, efforts are underway to improve
transparency through open data on contracts and payments. However, this
information is still only partially available, which limits large-scale
studies of the impact of transformative agreements on the transition of
hybrid journals to full open access. By comparing open metadata from
hoaddata, an openly available dataset on hybrid open access based on
metadata from Crossref, OpenAlex and the cOAlition S Journal Checker
Tool, with the established bibliometric databases Scopus and Web of
Science, this study aims to provide a robust estimate of the impact of
transformative agreements on hybrid open access. This comparative
approach informs bibliometric research and practice by highlighting the
strengths and weaknesses of each data source individually and
collectively.

Since their initial proposal by the Max Planck Digital Library (Schimmer
et al., 2015), transformative agreements have led to rapid growth in the
number of negotiated agreements and open access articles (Jahn, 2025;
McCabe \& Mueller-Langer, 2024; Rothfritz et al., 2024). Before, only a
small proportion of articles in hybrid journals were openly available
(Björk, 2012; Laakso \& Björk, 2016; Piwowar et al., 2018), with
publications from only a few European countries with dedicated funding
policies and the high-energy physics SCOAP\^{}3 consortium standing out
(Huang et al., 2020; Kohls \& Mele, 2018; Pinfield, 2015;
Robinson-Garcia et al., 2020). Related coordination activities, in
particular through the OA2020 and ESAC (Efficiency and Standards for
Article Charges) initiatives, have also led to improved workflows
between library consortia and publishers (Campbell et al., 2022;
Geschuhn \& Stone, 2017). However, critics have raised concerns about
these agreements, particularly regarding market concentration (Butler et
al., 2023; Shu \& Larivière, 2023), the failure of most hybrid journals
to convert to full open access (Kiley, 2024; Momeni et al., 2021),
reduced competition and incentives (McCabe \& Mueller-Langer, 2024;
Schmal, 2024), and the continuation of publication fees that widen gaps
between well-resourced and under-resourced institutions (Babini et al.,
2022; Ross-Hellauer et al., 2022)

While consortia evaluations confirm the growth of open access through
transformative agreements and acknowledge increasing coordination, they
reach mixed policy conclusions. For instance, the British Jisc
evaluation (Brayman et al., 2024) found that the consortium's
transformative agreements had a significant impact on national open
access growth but a limited effect on the global open access transition,
suggesting a need to reevaluate their strategy. Similar policy
reconsiderations emerged in Norway and Sweden (Holden et al., 2023;
Widding, 2024). Meanwhile, the funder initiative cOAlition S recommended
ending financial support for these agreements in 2024, but still
considers hybrid open access to be compliant with funders' open access
policies. Conversely, the German DEAL consortium extended agreements
with major publishers until 2028, while also outside of Europe, more
transformative agreements are being put in place (for instance in Latin
America, as described in Muñoz-Vélez et al. (2024)). Responding to
criticism, the Max Planck Digital Library recently outlined strategies
to evolve the original model, focusing on achieving fully open access
while controlling costs and addressing identified weaknesses (Dér, 2025)

Despite controversies, transformative agreements can benefit the
landscape of open research information as library consortia negotiate
for comprehensive open metadata, helping avoid data analytics from
becoming a commodity controlled by publishers (Aspesi \& Brand, 2020;
McCabe \& Mueller-Langer, 2024). For instance, ESAC recommends using
Crossref (Hendricks et al., 2020) to share metadata supporting workflows
including open access licenses and funding information (Geschuhn \&
Stone, 2017). Simultaneously, the bibliometric community pushes for more
open metadata through this publisher-driven DOI registration platform
and continuously monitors the progress (N. J. van Eck \& Waltman, 2022).

Many open scholarly data services that build on Crossref have benefited
from these developments. A prominent example is OpenAlex, which has
gained attention for its inclusivity, openness and comprehensive
coverage of publication metadata (Priem et al., 2022). In a recent
evaluation, Céspedes et al. (2025) reviewed the current evidence base on
the strengths and weaknesses of OpenAlex. They highlight both its
increasing use in bibliometric research and its adoption by research
organisations. Comparative studies have generally found OpenAlex to be a
reliable alternative to commercial databases (Alperin et al., 2024;
Culbert et al., 2025), although data quality issues occurred for key
metadata fields required for open access analysis, including evidence of
open access status itself (Jahn et al., 2023; Simard et al., 2024), as
well as authorships (Culbert et al., 2025), affiliations (Zhang et al.,
2024), author roles (Jahn, 2025) and document types used to distinguish
between different article types (Haupka et al., 2024). However, these
studies highlight the rapid improvements and new possibilities of
combining open metadata and the curation efforts of OpenAlex in
conjunction with the growing awareness of the importance of open
metadata, as demonstrated by the Barcelona Declaration on Open Research
Information.

Consequently, there is growing interest in using open bibliometric
databases such as Crossref and OpenAlex to assess the impact of
transformative agreements. Similar to established proprietary
bibliometric databases such as Scopus and Web of Science, these data
sources lack specific data points, particularly about open access
payments, which are typically not shared (Haustein et al., 2024).
However, bibliometric data can be enriched with open metadata about
journals and institutions involved in transformative agreements (Jahn,
2025). Jonge et al. (2025) demonstrated this potential by combining data
from the cOAlition S Journal Checker tool and OpenAlex, achieving 89\%
accuracy when validated against Dutch research information. While open
data approaches show promise, understanding how these results compare
when the same methodology is applied to proprietary databases such as
Web of Science and Scopus remains unexplored.

This study compares open and proprietary data sources for evaluating
transformative agreements. It adapts the approach from Jahn (2025),
which combines data from the cOAlition S Journal Checker tool, Crossref
and OpenAlex, and applies the same methods to Web of Science and Scopus
data. This comparative approach allows the evaluation of different
bibliometric data sources to monitor transformative agreements, while
assessing their respective strengths and limitations. Specifically, this
study compares the indexing of more than 13,000 hybrid journals across
these sources, and examines whether consistent results can be obtained
in terms of open access adoption between 2019 and 2023, analysed by
publishers and countries. By comparing results across these data
sources, this large-scale study provides a more robust understanding of
the impact of transformative agreements on the transition to full open
access. It also reveals the comparative strengths and weaknesses of
different bibliometric databases in measuring scholarly publishing's
shift toward openness.

@@

hoaddata followed good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex.\footnote{\url{https://subugoe.github.io/scholcomm_analytics/data.html}}
The workflow has run regularly to fetch updates from these data sources
since 2022. The package version used in this study is 0.3, containing
data from the Crossref 2024-08 dump provided to Crossref Metadata Plus
subscribers and the OpenAlex 2024-08-29 monthly dump. It covers
agreements collected between July 2021 to July 2024 from the JCT. This
version including the computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is a selective database with a
focus on base research (Stahlschmidt \& Stephen, 2022; Visser et al.,
2021). The collections considered in this study were the Science
Citation Index Expanded (SCIE), Social Sciences Citation Index (SSCI)
and Arts \& Humanities Citation Index (AHCI).

These collections provide important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall (Piwowar et al.,
2018), the same provider as OpenAlex. However, Web of Science lacks
information about journals and articles under transformative agreements.

For programmatic access to article-level data, this study used the
database of the Kompetenznetzwerk Bibliometrie (KB) in Germany. The KB
processes raw XML data provided by Clarivate Analytics, which are
ingested into an in-house PostgreSQL database under a uniform schema. To
support reproducibility, KB maintains annual snapshots of the database.
Accordingly, this study used the annual snapshot from April 2024
(wos\_b\_202404), which is considered to cover almost the entire
previous publication year (Schmidt et al., 2024).

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to Web of Science, Scopus is selective with regard to the
journals it indexes. However, its journal coverage is much broader than
that of the Web of Science collections considered in this study, as it
indexes a wider range of applied research journals (Singh et al., 2021;
Stahlschmidt \& Stephen, 2022; Visser et al., 2021). With detailed
metadata about article types, open access status information derived
from Unpaywall, author roles, and disambiguated affiliations, Scopus
also contains important data to assess open access uptake, although no
direct information regarding transformative agreements was available at
the time of the study.

This study used the Scopus annual snapshot from April 2024 as provided
by KB (scp\_b\_202404). The same KB curation effort as that for Web of
Science was applied to Scopus (Schmidt et al., 2024).

@@

\subsubsection{Determining the hybrid journal publication
volume}\label{determining-the-hybrid-journal-publication-volume}

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L. To
identify hybrid journals, a comprehensive exclusion of fully open access
journals was performed using multiple journal lists including the
Directory of Open Access Journals (DOAJ). The resulting hybrid journal
data were made available via hoaddata and were used to independently
determine the publication volume for each database.

hoaddata relies on Crossref to obtain journal publication volume and
open access status through Creative Commons (CC) licence information
relative to the published version (``version of record''). The article
metadata included DOIs, publication dates, open access information, and
author roles and affiliations. Publication years were determined using
the earliest known date of publication in a journal. In hoaddata, this
corresponded to Crossref's issued date. For Web of Science and Scopus,
the earliest publication date was used where available, with Scopus
dates specifically determined by the KB through version tracking of the
raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews (Borrego et al., 2021). Because of limited information on these
document types in open scholarly data (Haupka et al., 2024), hoaddata
used an extended version of Unpaywall's paratext recognition approach to
exclude non-scholarly content (Piwowar et al., 2018). To exclude
conference supplements, which are often not covered by transformative
agreements, only articles published in regular issues, indicated by
numerical pagination, were considered. For Web of Science and Scopus,
their established and mainly accurate document type classifications
(Donner, 2017; Maisano et al., 2025) were used to identify original
research articles and reviews, referred to as original articles
throughout this study.

@@

Articles in hybrid journals were considered open access when publishers
made them freely available under a CC license on their platforms. While
hoaddata obtained this information from Crossref license metadata, Web
of Science and Scopus relied on Unpaywall as evidence source. Unpaywall
also uses Crossref license metadata, but supplements them by parsing
publisher websites directly, addressing cases where publishers do not
provide machine-readable CC license information (Piwowar et al., 2018).
This additional parsing remains necessary despite transformative
agreement workflows recommending the deposition of CC licenses during
DOI registration (Geschuhn \& Stone, 2017). Both Web of
Science\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}
and Scopus\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}}
defined hybrid open access consistently as content available under CC
licenses on publisher platforms, distinguishing it from bronze open
access that lack such explicit license information, or use
publisher-specific licenses (Piwowar et al., 2018).

@@

Author affiliations were retrieved for both the first and, if available,
the corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To improve data
retrieval, JCT institution data were enriched with ROR-IDs from
associated institutions, such as university hospitals or institutes of
large research organisations such as the Max Planck Society, according
to OpenAlex's institution entity. Database-specific affiliation
identifiers were used to handle different address variants: ROR-IDs from
OpenAlex for hoaddata, affiliation enhanced names for Web of Science,
and Scopus Affiliation Identifier. Additionally, ISO country codes were
retrieved for each author's address to compile country-level statistics.

Because neither Web of Science nor Scopus supported ROR-ID at the time
of data retrieval, the institution identifier used by the JCT, a
two-step matching process was implemented to harmonise the affiliation
data. First, 2,782,540 articles from 6,457 institutions with ROR-IDs in
the JCT data since 2017 (according to hoaddata) were processed to map
the first authors' ROR-IDs to corresponding proprietary affiliation
identifiers in Web of Science and Scopus using DOI matching. Then, an
algorithm selected the most frequent ROR ID and proprietary identifier
pairs to handle multiple affiliations and organisational hierarchy
differences.

This process linked 6,375 ROR-IDs to 4,894 Scopus Affiliation IDs and
6,034 ROR-IDs to 2,422 enhanced affiliation strings in Web of Science.
Quality evaluation through random sampling of 50 pairs revealed an error
rate of 22\% for Web of Science (11 mismatches) and 6\% for Scopus
(three mismatches). Upon inspection, these mismatches primarily occurred
with less-represented institutions having only a few publications,
introduced through multiple affiliations of single authors. The
difference between databases suggests that Scopus' affiliation control
aligns more closely with ROR than with Web of Science.

@@

Based on these compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain ROR-IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included
matching of both journals and participating institutions according to
the JCT. The matching also considered the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (Jonge et al., 2025), applied to publications
funded by the Dutch Research Council (NWO) and validated against
internal invoicing data, confirmed that such matching can accurately
identify most articles under transformative agreements.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at the country and journal
levels by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of coverage between 2019 and 2023. It shows that the majority
of hybrid journals published at least one original research article or
review, collectively marked as ``original article'' in the following,
during the five-year period. These journals have provided the basis for
the subsequent calculation of article-level indicators.

@@

While hoaddata only covered articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, because Unpaywall only collects information on open access
status for articles with DOI. A closer look at original articles with
affiliation data, in line with related research (Zhang et al., 2024),
reveals a lack of affiliation data in the case of OpenAlex, the
affiliation data source used by hoaddata, compared to Web of Science and
Scopus. In particular, only approximately two-thirds of the articles
examined provided corresponding author affiliations. The proportion for
first authors was 89\%. At the time of writing, OpenAlex disclosed
limited coverage of corresponding authorship data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

@@

TTidyverse tools (Wickham et al., 2019) for the R programming language
(R Core Team, 2024) were used throughout this largely automated data
collection and analysis process. Rank correlations were calculated using
the Hmisc package (Harrell Jr, 2003). Source code analysis, including
all queries used to obtain the data, is available on GitHub
(\url{https://github.com/njahn82/hoa_validation}). An interactive
supplement exploring correlations between the data sources examined by
country and publisher is available via HuggingFace Spaces:
\url{https://huggingface.co/spaces/najkoja/hoa_replication}.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals and articles are visualised as an UpSet graph (Krassowski,
2020; Lex et al., 2014). The analysis included hybrid journals that
published at least one open access article between 2019 and 2023, based
on open access status information from each database. Only original
articles were included in the analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of the hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were found only in Scopus. Of these, 354 were available in Web of
Science. Upon inspection, this group of hybrid journals exclusively
covered in proprietary data sources mainly represented hybrid journals
for which no open access evidence could be retrieved from Crossref, the
open access evidence source for hoaddata.
